---
title: "Tiny Project - COVID-19 Health Statistics Analysis"
author: "Tanmay Gupta"
date: "2026-09-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This report analyzes the spread of COVID-19 across Indian states using daily
reported data on confirmed cases, cured patients, and deaths. The dataset used
is `covid_19_india.csv`, sourced from Kaggle.

---

## 1. Data Importing

```{r}
# Read the CSV file into R (must be in the same folder as this .Rmd file)
covid_data <- read.csv("covid_19_india.csv")

# Preview the first few rows
head(covid_data)
```

**Explanation:** `read.csv()` loads the dataset into a data frame called
`covid_data`. `head()` shows the first 6 rows so we can confirm it loaded
correctly.

---

## 2. Data Structure and Overview

```{r}
str(covid_data)
```

```{r}
summary(covid_data)
```

**Explanation:** `str()` shows the data type of each column (text, number,
etc.). `summary()` gives quick statistics (min, max, average) for numeric
columns like Confirmed, Cured, and Deaths.

---

## 3. Data Preprocessing and Cleaning

```{r}
# Check for missing values in each column
colSums(is.na(covid_data))
```

```{r}
# Convert Date column from text to proper Date format
covid_data$Date <- as.Date(covid_data$Date, format = "%Y-%m-%d")

# Confirm conversion worked
class(covid_data$Date)
```

```{r}
# Check for duplicate rows
sum(duplicated(covid_data))
```

**Explanation:** `colSums(is.na(...))` counts missing values per column.
Converting `Date` to a proper date type lets R understand chronological order
(needed for the trend chart later). We also check for duplicate rows to
ensure data quality.

**Important note on this dataset:** each row stores a *cumulative running
total* for a state as of that date (not new daily cases). This means we must
NOT simply add up every row later — that would multiply the same growing
numbers across ~500 days and give impossible totals. Instead, for any
"grand total," we will use only the **most recent date's** rows, since those
already represent the final cumulative count for each state.

---

## 4. Data Analysis and Statistical Exploration

```{r}
# Isolate only the most recent date's data (correct final totals per state)
latest_date <- max(covid_data$Date)
latest_data <- covid_data[covid_data$Date == latest_date, ]

latest_date
```

```{r}
# Overall totals across India (as of the latest date)
total_confirmed <- sum(latest_data$Confirmed)
total_cured <- sum(latest_data$Cured)
total_deaths <- sum(latest_data$Deaths)

total_confirmed
total_cured
total_deaths
```

```{r}
# Which states have the highest confirmed cases (as of latest date)?
state_totals <- latest_data[order(-latest_data$Confirmed),
                             c("State.UnionTerritory", "Confirmed")]
head(state_totals, 10)
```

```{r}
# Central tendency and spread of confirmed cases across states (latest date)
median(latest_data$Confirmed)
var(latest_data$Confirmed)
sd(latest_data$Confirmed)
```

```{r}
# Frequency count: number of days each state has reported data
state_freq <- table(covid_data$State.UnionTerritory)
head(state_freq)
```

```{r}
# Correlation between Confirmed cases and Deaths (using full daily data)
cor(covid_data$Confirmed, covid_data$Deaths)
```

```{r}
# Correlation between Confirmed cases and Cured (using full daily data)
cor(covid_data$Confirmed, covid_data$Cured)
```

**Explanation:** We use `latest_data` (a single date's snapshot) for grand
totals and state rankings, since the data is cumulative. `median()`, `var()`,
and `sd()` show how spread out the case counts are across states — a large
standard deviation means some states had far more cases than others.
`table()` counts how many days of data exist per state, useful for checking
data completeness. For correlation, using the full daily dataset is fine and
standard practice, because correlation measures the *relationship* between
two columns' growth pattern, not an absolute total.

---

## 5. Graphs and Plots

```{r}
# Bar chart: Top 10 states by confirmed cases
top10 <- head(state_totals, 10)

barplot(
  top10$Confirmed,
  names.arg = top10$State.UnionTerritory,
  main = "Top 10 States by Confirmed COVID-19 Cases",
  xlab = "State",
  ylab = "Confirmed Cases",
  las = 2,
  col = "steelblue"
)
```

```{r}
# Pie chart: overall outcome split (Cured vs Deaths vs Active)
active <- total_confirmed - total_cured - total_deaths

outcome <- c(Cured = total_cured, Deaths = total_deaths, Active = active)

pie(
  outcome,
  main = "Overall Case Outcome Distribution",
  labels = paste(names(outcome), outcome)
)
```

```{r}
# Line trend: total confirmed cases over time (all states combined)
daily_totals <- aggregate(Confirmed ~ Date, data = covid_data, sum)

plot(
  daily_totals$Date,
  daily_totals$Confirmed,
  type = "l",
  main = "COVID-19 Confirmed Cases Over Time (All India)",
  xlab = "Date",
  ylab = "Total Confirmed Cases",
  col = "darkred"
)
```

```{r}
# Scatter plot: Confirmed vs Deaths relationship
plot(
  covid_data$Confirmed,
  covid_data$Deaths,
  main = "Confirmed Cases vs Deaths",
  xlab = "Confirmed Cases",
  ylab = "Deaths",
  pch = 19,
  col = "darkgreen"
)
```

**Explanation:** The bar chart highlights the worst-affected states as of
the most recent date. The pie chart shows the overall proportion of
patients cured, deceased, or still active. The line plot shows how the
national case count grew over time. The scatter plot visualizes the
relationship between case count and deaths across all states and dates.

---

## 6. Findings and Conclusion

- The state of **Maharashtra** recorded the highest number of confirmed
  cases (6,363,442), suggesting it was the most affected region in the
  country, followed by Kerala and Karnataka.
- The correlation between Confirmed cases and Deaths was **0.918**,
  indicating a **strong** positive relationship — meaning states with more
  cases generally also recorded proportionally more deaths.
- The correlation between Confirmed cases and Cured was **0.998**, an
  even **stronger** relationship, showing recoveries closely tracked the
  rise in confirmed cases over time.
- The trend line shows cases **rose sharply** in two distinct phases —
  a steep climb during mid-to-late 2020, followed by an even sharper spike
  in **April-May 2021**, reflecting India's second COVID-19 wave.
- Overall, out of all confirmed cases (32,036,511 total), **97.45%** were
  cured, and **1.34%** resulted in death, showing a **high** recovery rate.

**Conclusion:** This analysis of India's COVID-19 data shows how the
pandemic spread unevenly across states, with Maharashtra, Kerala, and
Karnataka bearing a disproportionate share of cases. The strong correlation
between cases and deaths/recoveries highlights the scale of the healthcare
response required during the outbreak, while the high overall recovery rate
of over 97% reflects the effectiveness of treatment and containment efforts
over time.